---
title: "downstream_model11"
author: "Christina Morgenstern"
date: "6/7/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```

# Downstream analysis of MOFA-PDAC model 11

```{r}
library(ggplot2)
library(cowplot)
# for GSEA analysis
library(msigdbr)

# MOFA
library(MOFAdata)
library(MOFAdata)
```

```{r}
model <- load_model('/Users/christina/01_Files/19_Master studies/02_Course work/13_Capstone/data/data/models_final/MOFA_PAAD_11.hdf5')
```

```{r}
plot_data_overview(model)
```

## Add metadata to the model

### Import clincial metadata

```{r}
# clinical metadata
clinical_metatdata <- read.csv("/Users/christina/01_Files/19_Master studies/02_Course work/13_Capstone/data/data/csv_files/clinical_metadata.csv")
head(clinical_metatdata)
```

Prepare clinical metadata
```{r}
#transpose data frame (samples rows and features columns)
metadata <- setNames(data.frame(t(clinical_metatdata[,-1])), clinical_metatdata[,1])
metadata <- rownames_to_column(metadata, var = "sample")

# convert non-numeric columns to numeric
metadata[, c(2, 8, 14, 15, 16)] <- sapply(metadata[, c(2, 8, 14, 15, 16)], as.numeric)
head(metadata)
```


```{r}
samples_metadata(model) <- metadata 
```

## Variance decomposition

Quantify the amount of variance explained (R2) by each factor in each data modality.

```{r}
head(model@cache$variance_explained$r2_total)
```

```{r}
head(model@cache$variance_explained$r2_per_factor)
```
```{r}
plot_variance_explained(model, max_r2=20)
```
```{r}
plot_variance_explained(model, plot_total = T)
```

A good sanity check is to verify that the Factors are largely uncorrelated. In MOFA there are no orthogonality constraints such as in Principal Component Analysis, but if there is a lot of correlation between Factors this suggests a poor model fit
```{r}
plot_factor_cor(model)
```

```{r}
r2 <- model@cache$variance_explained$r2_per_factor[[1]]

r2.dt <- r2 %>%
  as.data.table %>% .[,factor:=as.factor(1:model@dimensions$K)] %>%
  melt(id.vars=c("factor"), variable.name="view", value.name = "r2") %>%
  .[,cum_r2:=cumsum(r2), by="view"]

ggline(r2.dt, x="factor", y="cum_r2", color="view") +
  labs(x="Factor number", y="Cumulative variance explained (%)") +
  theme(
    legend.title = element_blank(), 
    legend.position = "top",
    axis.text = element_text(size=rel(0.8))
  )
```

```{r}
pathologic_stage.colors <- c(
  "nan" = "grey",
  "stagei" = "#66C2A5", 
  "stageii" = "#8DA0CB",
  "stageiii" = "#E78AC3",
  "stageiv" = "#FC8D62"
)
```


```{r}
plot_factor(model, 
  factor = 1, 
  color_by = "pathologic_stage", 
  dot_size = 4,
  dodge = TRUE,
  stroke = 0.4,
  add_violin = T,
  add_boxplot = T
) +
  scale_fill_manual(values=pathologic_stage.colors) +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank()
  )
```

```{r}
plot_factor(model, 
  factor = 2, 
  color_by = "pathologic_stage", 
  dot_size = 4,
  dodge = TRUE,
  stroke = 0.4,
  add_violin = T,
  add_boxplot = T
) +
  scale_fill_manual(values=pathologic_stage.colors) +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank()
  )
```

```{r}
plot_factor(model, 
  factor = 3, 
  color_by = "pathologic_stage", 
  dot_size = 4,
  dodge = TRUE,
  stroke = 0.4,
  add_violin = T,
  add_boxplot = T
) +
  scale_fill_manual(values=pathologic_stage.colors) +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank()
  )
```

```{r}
status.colors <- c(
  "nan" = "grey",
  "0" = "#66C2A5", 
  "1" = "#8DA0CB"
)
```


```{r}
plot_factor(model, 
  factor = 1, 
  color_by = "status", 
  dot_size = 4,
  dodge = TRUE,
  stroke = 0.4,
  add_violin = T,
  add_boxplot = T
) +
  scale_fill_manual(values=status.colors) +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank()
  )
```

```{r}
plot_factor(model, 
  factor = 2, 
  color_by = "status", 
  dot_size = 4,
  dodge = TRUE,
  stroke = 0.4,
  add_violin = T,
  add_boxplot = T
) +
  scale_fill_manual(values=status.colors) +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank()
  )
```

```{r}
plot_factor(model, 
  factor = 3, 
  color_by = "status", 
  dot_size = 4,
  dodge = TRUE,
  stroke = 0.4,
  add_violin = T,
  add_boxplot = T
) +
  scale_fill_manual(values=status.colors) +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank()
  )
```


```{r}
correlate_factors_with_covariates(model, 
  covariates = c("pathologic_stage", "pathology_T_stage", "pathology_N_stage", "pathology_M_stage", "gender", "radiation_therapy", "residual_tumor", "race", "ethnicity", "overall_survival", "status", "histological_type", "residual_tumor", "radiation_therapy"),
  plot = "r",  # use "log_pval" to plot log p-values 
)
```



```{r}
plot_factor(model, 
            factor = 1:3,
            color_by = "overall_survival",
            group_by = "gender",
            add_violin = TRUE
            )
```
```{r}
plot_factors(model, 
  factors = 1:3,
  color_by = "overall_survival",
  shape_by = "gender"
)
```

### Association analysis

```{r}
correlate_factors_with_covariates(model, covariates = c("pathologic_stage", "pathology_T_stage", "pathology_N_stage", "pathology_M_stage", "gender", "radiation_therapy", "residual_tumor", "race", "ethnicity", "overall_survival"),
                                  plot = "log_pval")
```

### Characterisation of factors

Plot feature weights

```{r}
plot_top_weights(model,
                  view = "mRNA",
                  factor = 1,
                  nfeatures = 10,
                  scale = T
                  )
```

```{r}
plot_top_weights(model,
                  view = "mutation",
                  factor = 1,
                  nfeatures = 10,
                  scale = T
                  )
```

```{r}
plot_factor(mofa_object, 
  factors = 1, 
  color_by = "KRAS_mutation",
  add_violin = TRUE,
  dodge = TRUE
)

```


```{r}
plot_top_weights(model,
                  view = "methylation",
                  factor = 1,
                  nfeatures = 10,
                  scale = T
                  )
```

```{r}
plot_top_weights(model,
                  view = "SCNV",
                  factor = 1,
                  nfeatures = 10,
                  scale = T
                  )
```

```{r}
plot_top_weights(model,
                  view = "microRNA",
                  factor = 1,
                  nfeatures = 10,
                  scale = T
                  )
```

```{r}
plot_top_weights(model,
                  view = "protein",
                  factor = 1,
                  nfeatures = 10,
                  scale = T
                  )
```

```{r}
plot_factor(model, 
  factors = 1, 
  color_by = "KRAS_mutation",
  dodge = TRUE,
  add_violin = TRUE
)
```

```{r}
p <- plot_factors(model, 
  factors = c(1,3), 
  color_by = "KRAS_mutation",
  shape_by = "TP53_mutation",
  dot_size = 2.5,
  show_missing = T
)

p <- p + 
  geom_hline(yintercept=-1, linetype="dashed") +
  geom_vline(xintercept=(-0.5), linetype="dashed")

print(p)
```


Characterization of Factor 2

```{r}
plot_top_weights(model,
                  view = "mRNA",
                  factor = 2,
                  nfeatures = 10,
                  scale = T
)
```


```{r}
plot_top_weights(model,
                 view = "mutation",
                 factors = 2,
                 nfeatures = 10)
```


```{r}
plot_top_weights(model,
                  view = "methylation",
                  factor = 2,
                  nfeatures = 10,
                  scale = T
                  )
```

```{r}
plot_top_weights(model,
                  view = "SCNV",
                  factor = 2,
                  nfeatures = 10,
                  scale = T
                  )
```

```{r}
plot_top_weights(model,
                  view = "microRNA",
                  factor = 2,
                  nfeatures = 10,
                  scale = T
                  )
```

```{r}
plot_top_weights(model,
                  view = "protein",
                  factor = 2,
                  nfeatures = 10,
                  scale = T
                  )
```


Factor 3

```{r}
plot_top_weights(model,
                  view = "mRNA",
                  factor = 3,
                  nfeatures = 10,
                  scale = T
)
```

```{r}
plot_top_weights(model,
                  view = "mutation",
                  factor = 3,
                  nfeatures = 10,
                  scale = T
)
```


```{r}
plot_top_weights(model,
                  view = "methylation",
                  factor = 3,
                  nfeatures = 10,
                  scale = T
                  )
```
```{r}
plot_top_weights(model,
                  view = "microRNA",
                  factor = 3,
                  nfeatures = 10,
                  scale = T
                  )
```

```{r}
plot_top_weights(model,
                  view = "SCNV",
                  factor = 3,
                  nfeatures = 10,
                  scale = T
                  )
```


```{r}
plot_top_weights(model,
                  view = "protein",
                  factor = 3,
                  nfeatures = 10,
                  scale = T
                  )
```

```{r}
plot_data_scatter(model, 
  view = "mRNA",
  factor = 1,  
  features = 4,
  sign = "negative",
  color_by = "CDKN2A_SCNV"
) + labs(y="RNA expression")
```

```{r}
plot_data_scatter(model, 
  view = "mRNA",
  factor = 1,  
  features = 4,
  sign = "negative",
  color_by = "C9orf53_SCNV"
) + labs(y="RNA expression")
```
```{r}
plot_data_scatter(model, 
  view = "methylation",
  factor = 2,  
  features = 4,
  sign = "negative",
  color_by = "KRAS_mutation"
) + labs(y="RNA expression")
```


```{r}
plot_data_scatter(model, 
  view = "mRNA",
  factor = 1,  
  features = 4,
  sign = "negtive",
  color_by = "CELSR1_methylation"
) + labs(y="RNA expression")
```
```{r}
plot_data_scatter(model, 
  view = "methylation",
  factor = 1,  
  features = 4,
  sign = "positive",
  color_by = "CEACAM5_mRNA"
) + labs(y="RNA expression")
```

```{r}
plot_data_scatter(model, 
  view = "methylation",
  factor = 1,  
  features = 4,
  sign = "positive",
  color_by = "hsa-mir-199a-2"
) + labs(y="RNA expression")
```


### Visualization of patterns in input data

```{r}
plot_data_heatmap(model,
  view = "mRNA",         # view of interest
  factor = 1,             # factor of interest
  features = 10,          # number of features to plot (they are selected by weight)
  
  # extra arguments that are passed to the `pheatmap` function
  cluster_rows = TRUE, cluster_cols = FALSE,
  show_rownames = TRUE, show_colnames = FALSE,
  annotation_samples =  c("pathologic_stage", "status", "overall_survival"),
  denoise = TRUE
)
```

```{r}
plot_data_heatmap(model,
  view = "mRNA",         # view of interest
  factor = 1,             # factor of interest
  features = 10,          # number of features to plot (they are selected by weight)
  
  # extra arguments that are passed to the `pheatmap` function
  cluster_rows = TRUE, cluster_cols = FALSE,
  show_rownames = TRUE, show_colnames = FALSE,
  annotation_samples = "pathology_T_stage",
  denoise = TRUE
)
```


```{r}
plot_data_heatmap(model,
  view = "mRNA",         # view of interest
  factor = 2,             # factor of interest
  features = 10,          # number of features to plot (they are selected by weight)
  
  # extra arguments that are passed to the `pheatmap` function
  cluster_rows = TRUE, cluster_cols = FALSE,
  show_rownames = TRUE, show_colnames = FALSE,
  annotation_samples = "pathologic_stage",
  denoise = TRUE
)
```
```{r}
plot_data_heatmap(model,
  view = "mRNA",         # view of interest
  factor = 3,             # factor of interest
  features = 10,          # number of features to plot (they are selected by weight)
  
  # extra arguments that are passed to the `pheatmap` function
  cluster_rows = TRUE, cluster_cols = FALSE,
  show_rownames = TRUE, show_colnames = FALSE,
  annotation_samples = c("pathologic_stage", "status", "overall_survival"),
  denoise = TRUE
)
```


```{r}
plot_data_heatmap(model,
  view = "microRNA",         # view of interest
  factor = 1,             # factor of interest
  features = 10,          # number of features to plot (they are selected by weight)
  
  # extra arguments that are passed to the `pheatmap` function
  cluster_rows = TRUE, cluster_cols = FALSE,
  show_rownames = TRUE, show_colnames = FALSE,
  annotation_samples = c("pathologic_stage", "status", "overall_survival"),
  denoise = TRUE
)
```

```{r}
plot_data_heatmap(model,
  view = "microRNA",         # view of interest
  factor = 3,             # factor of interest
  features = 10,          # number of features to plot (they are selected by weight)
  
  # extra arguments that are passed to the `pheatmap` function
  cluster_rows = TRUE, cluster_cols = FALSE,
  show_rownames = TRUE, show_colnames = FALSE,
  annotation_samples = c("pathologic_stage", "status", "overall_survival"),
  denoise = TRUE
)
```

```{r}
plot_data_heatmap(model,
  view = "mutation",         # view of interest
  factor = 1,             # factor of interest
  features = 10,          # number of features to plot (they are selected by weight)
  
  # extra arguments that are passed to the `pheatmap` function
  cluster_rows = TRUE, cluster_cols = FALSE,
  show_rownames = TRUE, show_colnames = FALSE,
  annotation_samples = c("pathology_T_stage","overall_survival", "pathologic_stage")
)
```

```{r}
plot_data_heatmap(model,
  view = "mutation",         # view of interest
  factor = 2,             # factor of interest
  features = 10,          # number of features to plot (they are selected by weight)
  
  # extra arguments that are passed to the `pheatmap` function
  cluster_rows = TRUE, cluster_cols = FALSE,
  show_rownames = TRUE, show_colnames = FALSE,
  annotation_samples = c("pathology_T_stage","overall_survival", "pathologic_stage")
)
```

```{r}
plot_data_heatmap(model,
  view = "mutation",         # view of interest
  factor = 3,             # factor of interest
  features = 10,          # number of features to plot (they are selected by weight)
  
  # extra arguments that are passed to the `pheatmap` function
  cluster_rows = TRUE, cluster_cols = FALSE,
  show_rownames = TRUE, show_colnames = FALSE,
  annotation_samples = c("pathology_T_stage","overall_survival", "pathologic_stage")
)
```

```{r}
plot_data_heatmap(model,
  view = "SCNV",         # view of interest
  factor = 1,             # factor of interest
  features = 10,          # number of features to plot (they are selected by weight)
  
  # extra arguments that are passed to the `pheatmap` function
  cluster_rows = TRUE, cluster_cols = FALSE,
  show_rownames = TRUE, show_colnames = FALSE,
  annotation_samples = c("pathology_T_stage","overall_survival", "pathologic_stage"),
  denoise = TRUE
)
```

```{r}
plot_data_heatmap(model,
  view = "SCNV",         # view of interest
  factor = 2,             # factor of interest
  features = 10,          # number of features to plot (they are selected by weight)
  
  # extra arguments that are passed to the `pheatmap` function
  cluster_rows = TRUE, cluster_cols = FALSE,
  show_rownames = TRUE, show_colnames = FALSE,
  annotation_samples = c("pathology_T_stage","overall_survival", "pathologic_stage")
)
```
```{r}
plot_data_heatmap(model,
  view = "SCNV",         # view of interest
  factor = 3,             # factor of interest
  features = 10,          # number of features to plot (they are selected by weight)
  
  # extra arguments that are passed to the `pheatmap` function
  cluster_rows = TRUE, cluster_cols = FALSE,
  show_rownames = TRUE, show_colnames = FALSE,
  annotation_samples = c("pathology_T_stage","overall_survival", "pathologic_stage")
)
```


```{r}
plot_data_heatmap(model,
  view = "methylation",         # view of interest
  factor = 1,             # factor of interest
  features = 10,          # number of features to plot (they are selected by weight)
  
  # extra arguments that are passed to the `pheatmap` function
  cluster_rows = TRUE, cluster_cols = FALSE,
  show_rownames = TRUE, show_colnames = FALSE,
  annotation_samples = c("pathologic_stage", "status", "overall_survival"),
  denoise = TRUE
)
```


```{r}
plot_data_heatmap(model,
  view = "methylation",         # view of interest
  factor = 2,             # factor of interest
  features = 10,          # number of features to plot (they are selected by weight)
  
  # extra arguments that are passed to the `pheatmap` function
  cluster_rows = TRUE, cluster_cols = FALSE,
  show_rownames = TRUE, show_colnames = FALSE,
  annotation_samples = c("pathology_T_stage","overall_survival", "pathologic_stage")
)
```


```{r}
plot_data_heatmap(model,
  view = "methylation",         # view of interest
  factor = 3,             # factor of interest
  features = 10,          # number of features to plot (they are selected by weight)
  
  # extra arguments that are passed to the `pheatmap` function
  cluster_rows = TRUE, cluster_cols = FALSE,
  show_rownames = TRUE, show_colnames = FALSE,
  annotation_samples = c("pathologic_stage", "status", "overall_survival"),
  denoise = TRUE
)
```

```{r}
plot_data_heatmap(model,
  view = "protein",         # view of interest
  factor = 1,             # factor of interest
  features = 10,          # number of features to plot (they are selected by weight)
  
  # extra arguments that are passed to the `pheatmap` function
  cluster_rows = TRUE, cluster_cols = FALSE,
  show_rownames = TRUE, show_colnames = FALSE,
  annotation_samples = c("pathology_T_stage","overall_survival", "pathologic_stage")
)
```

```{r}
plot_data_heatmap(model,
  view = "protein",         # view of interest
  factor = 2,             # factor of interest
  features = 10,          # number of features to plot (they are selected by weight)
  
  # extra arguments that are passed to the `pheatmap` function
  cluster_rows = TRUE, cluster_cols = FALSE,
  show_rownames = TRUE, show_colnames = FALSE,
  annotation_samples = c("pathology_T_stage","overall_survival", "pathologic_stage")
)
```

```{r}
plot_data_heatmap(model,
  view = "protein",         # view of interest
  factor = 3,             # factor of interest
  features = 10,          # number of features to plot (they are selected by weight)
  
  # extra arguments that are passed to the `pheatmap` function
  cluster_rows = TRUE, cluster_cols = FALSE,
  show_rownames = TRUE, show_colnames = FALSE,
  annotation_samples = c("pathology_T_stage","overall_survival", "pathologic_stage")
)
```


```{r}
plot_data_scatter(model,
  view = "mRNA",         # view of interest
  factor = 1,             # factor of interest
  features = 5,           # number of features to plot (they are selected by weight)
  add_lm = TRUE,          # add linear regression
  color_by = "gender"
)
```

```{r}
plot_data_scatter(model,
  view = "SCNV",         # view of interest
  factor = 1,             # factor of interest
  features = 5,           # number of features to plot (they are selected by weight)
  add_lm = TRUE,          # add linear regression
  color_by = "gender"
)
```

```{r}
p <- plot_factors(mofa_object, 
  factors = c(1,3), 
  color_by = "C9orf53_SCNV",
  #shape_by = "C9orf53_SCNV",
  dot_size = 2.5,
  show_missing = T
)

p <- p + 
  geom_hline(yintercept=-1, linetype="dashed") +
  geom_vline(xintercept=(-0.5), linetype="dashed")

print(p)
```


```{r}
p2 <- plot_factors(mofa_object, 
  factors = c(1,2), 
  color_by = "CELSR1_methylation",
  #shape_by = "C9orf53_SCNV",
  dot_size = 2.5,
  show_missing = T
)

p2 <- p2 + 
  geom_hline(yintercept=-1, linetype="dashed") +
  geom_vline(xintercept=(-0.5), linetype="dashed")

print(p2)
```



## Non-linear dimensionality reduction

```{r}
set.seed(42)
model <- run_umap(model)
```

```{r}
plot_dimred(model,
            method = "UMAP",
            color_by = "gender")
```
```{r}
set.seed(42)
model <- run_tsne(model)
```

```{r}
plot_dimred(model,
            method = "TSNE",
            color_by = "gender")
```

### Imputation of missing values

```{r}
model <- impute(model)
```

```{r}
model@imputed_data
```


```{r}
genes <- list("CELF4_mRNA","GRIA2_mRNA")

genes %>% map(~ plot_factors(model, 
  factors = c(1,3), 
  color_by = ., 
  scale = T,
  legend = F
)) %>% cowplot::plot_grid(plotlist=., nrow=1)
```


```{r}
genes <- list("CD79A_mRNA","CXCL13_mRNA")

genes %>% map(~ plot_factors(model, 
  factors = c(2,3), 
  color_by = ., 
  scale = T,
  legend = F
)) %>% cowplot::plot_grid(plotlist=., nrow=1)
```


